---
title: "Assignment1 : Activity monitor"
---

## Section 1 : Mean total steps per day

```{r setoptions,echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "asis", fig.path='figure/',fig.width=10, fig.height=8)
options(scipen = 1, digits = 2)
```

```{r printMeanMedian,, message=F, warning=F}
        library(data.table)

        wkout_df <- read.csv("activity.csv",sep=",",stringsAsFactors=FALSE,header=TRUE)
        wkout_dt <- data.table(wkout_df)

        # adding the total # of steps by date  
        sum_dt <- wkout_dt[,sum(steps),by="date"]
        setnames(sum_dt,"V1","total")

        # creating the histogram
        hist(sum_dt$total,main="Total Steps Hist", xlab="total steps")
        mean_total <- mean(sum_dt$total,na.rm=TRUE)
        median_total <- median(sum_dt$total,na.rm=TRUE)
```

The mean is `r mean_total`.

The median is `r median_total`.

## Section 2 : Average daily activity pattern
```{r printMaxmeanInterval}
        # calculating average steps across all days by time interval
        mean_dt <- wkout_dt[,mean(steps,na.rm=TRUE),by="interval"]
        setnames(mean_dt,"V1","mean")

        # plotting a line graph for average steps across days by time interval
        plot(mean_dt,type="l")
        axis(side=1, at=seq(0, 288,by=1))

        # fetching the time interval data with maximum average steps
        dt<- subset(mean_dt,mean_dt$mean==max(mean_dt$mean))
```

The 5-min interval on average that contains max # of steps is `r  dt$interval`.


## Section 3 : Imputing missing values
```{r imputing data, message=F, warning=F}
        library(Hmisc,quietly=TRUE)

        nullval <- sum(is.na(wkout_dt$steps))

        mean_dt <- wkout_dt[,mean(steps,na.rm=TRUE),by="date"]
        setnames(mean_dt,"V1","mean")
        mean_dt[is.na(mean_dt)] <- 0
        
        # merging the daily mean to all the rows
        m_dt <- merge(wkout_dt, mean_dt, by="date", all.x=TRUE)
        
        # imputing the NAs with the mean value
        for (i in seq(1, nrow(m_dt))) 
        {
                if (is.na(m_dt$steps[i])){
                        m_dt$steps[i] <- m_dt$mean[i]
                }
        }

        #adding the total # of steps by date of the imputed data
        sum_dt <- m_dt[,sum(steps,na.rm=TRUE),by="date"]
        setnames(sum_dt,"V1","total")
        print(nrow(sum_dt))

        # creating the histogram of the imputed data
        hist(sum_dt$total,main="Total Steps Hist (Imputed)", xlab="total steps (imputed)")
        imp_mean_total <- mean(sum_dt$total,na.rm=TRUE)
        imp_median_total <- median(sum_dt$total,na.rm=TRUE)

```

'NA' count is `r nullval`.

The imputed mean is `r imp_mean_total`.

The imputed median is `r imp_median_total`.

## Section 4 : Difference in activity patterns
```{r plottingByWeekend vs  data, message=F, warning=F}

        # populate weekday or weekend based on the date
        for (i in seq(1, nrow(wkout_dt))) 
        {
                day <- as.POSIXlt(as.Date(wkout_dt$date[i]))$wday
                if (day == 6 || day == 0){
                        wkout_dt$week[i] <- "weekend"
                }
                else{
                        wkout_dt$week[i] <- "weekday"
                }
        }
        
        # calculate average steps for both weekend and weekday subset
        dt<- subset(wkout_dt,wkout_dt$week=="weekday")
        weekday_dt <- dt[,mean(steps,na.rm=TRUE),by="interval"]
        setnames(weekday_dt,"V1","mean")
        for (i in seq(1, nrow(weekday_dt))) 
        {    
                weekday_dt$week[i] <- "weekday"
        }
        dt<- subset(wkout_dt,wkout_dt$week=="weekend")
        weekend_dt <- dt[,mean(steps,na.rm=TRUE),by="interval"]
        setnames(weekend_dt,"V1","mean")
        for (i in seq(1, nrow(weekend_dt))) 
        {    
                weekend_dt$week[i] <- "weekend"
        }
        
        # bind both weekend and weekday data
        mydt <- rbind(weekday_dt,weekend_dt)
        plot <- xyplot(mean ~ interval | week, xlab = "interval",data=mydt, layout = c(1, 2),
                       type = "l",main = "Average Steps by Interval")
        print(plot, position = c(0, .3, 1, .9), more = TRUE)


```